---
title: "Web Scraping"
author: "James Martinez"
date: "April 14, 2019 (Updated May 15)"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rvest)
library(tidyverse)
```

## Introduction

This report covers the web scraping of [Basketball-Reference](https://www.basketball-reference.com/) for 2018-2019 regular season player data, as well as the modification for easy use with the [current logistic regression model](http://rpubs.com/martij222/all-nba-ml) using the `rvest` and `tidyverse` packages.

## Web Scraping

2018-2019 player stats can be found [here](https://www.basketball-reference.com/leagues/NBA_2019_totals.html). `read_html` outputs an XML document from which we can extract the stats we're interested in.

```{r player data}
# Get appropriate webpage
url <- "https://www.basketball-reference.com/leagues/NBA_2019_totals.html"
webpage <- read_html(url)

# Get headers
headers <- webpage %>% 
  html_nodes("table#totals_stats > thead > tr > th") %>% 
  html_text %>% .[-1] # Omit "Rk" column

# Save stats into matrix
data <- webpage %>% 
  html_nodes("table#totals_stats > tbody > tr > td") %>% 
  html_text %>% 
  matrix(ncol = length(headers), byrow = TRUE)

# Save as a data frame and add column titles
data <- as.data.frame(data, stringsAsFactors = FALSE)
names(data) <- headers

# Check data
data
```

In order to prevent problems when implementing our model on this new data, we'll take all the columns that were present in the original data, rename them, and select them in the same order they were listed. We'll then transform them in the same way.

```{r}
# Save as correct data types and select variables to match Kaggle data
players.2019 <- data %>% mutate_at(vars(G:PTS), as.numeric) %>% select(c(Player, Tm, G, MP, PTS, ORB, DRB, TRB, AST, STL, BLK, TOV, PF, FGA, FG, FTA, FT, "3PA", "3P", Pos))

# Match column names to Kaggle data set
names(players.2019) <- c("playerID", "tmID", "GP", "minutes", "points", "oRebounds", "dRebounds", "rebounds", "assists", "steals", 
                         "blocks", "turnovers", "PF", "fgAttempted", "fgMade", "ftAttempted", "ftMade", "threeAttempted", "threeMade", "pos")

# Add position indicators
players.2019 <- players.2019 %>% 
  mutate(
    center = case_when(
      grepl("C", pos) ~ as.integer(1),
      TRUE ~ as.integer(0)),
    
    forward = case_when(
      grepl("F", pos) ~ as.integer(1),
      TRUE ~ as.integer(0)),
    
    guard = case_when(
      grepl("G", pos) ~ as.integer(1),
      TRUE ~ as.integer(0))) %>% 
  select(-pos)

players.2019
```

Before we add the extra features, we first need to add the `allstar` indicator.

```{r allstar}
# All-Star 2019 webpage
url <- "https://www.basketball-reference.com/allstar/NBA_2019.html"

# Extract names
allstar <- url %>%
  read_html %>% 
  html_nodes("table > tbody > tr > th > a") %>% 
  html_text

# Add allstar column
players.2019 <- players.2019 %>% 
  mutate(allstar = case_when(
    playerID %in% allstar ~ as.integer(1),
    TRUE ~ as.integer(0)
  ))
```

Now we can add all the new features.

```{r feature eng}
# Add per game stats
players.2019 <- players.2019 %>% 
  mutate(
    minutesPerGame = minutes / GP,
    pointsPerGame = points / GP,
    assistsPerGame = assists / GP,
    oReboundsPerGame = oRebounds / GP,
    dReboundsPerGame = dRebounds / GP,
    reboundsPerGame = rebounds / GP,
    stealsPerGame = steals / GP,
    blocksPerGame = blocks / GP,
    turnoversPerGame = turnovers / GP,
    fgAttemptedPerGame = fgAttempted / GP,
    fgMadePerGame = fgMade/ GP,
    ftAttemptedPerGame = ftAttempted / GP,
    ftMadePerGame = ftMade / GP,
    threeAttemptedPerGame = threeAttempted / GP,
    threeMadePerGame = threeMade / GP)

# GP Ratio, health, and stats compared to league 
players.2019 <- players.2019 %>% 
  mutate(
    GPRatio = GP / max(GP), 
    healthy = case_when(
      GPRatio >= 0.7 ~ as.integer(1),
      TRUE ~ as.integer(0)),
    lgPoints = points / sum(points),
    lgAssists = assists / sum(assists),
    lgRebounds = rebounds / sum(rebounds),
    lgDRebounds = case_when(
      dRebounds != 0 ~ dRebounds / sum(dRebounds),
      TRUE ~ 0),
    lgORebounds = case_when(
      oRebounds != 0 ~ oRebounds / sum(oRebounds),
      TRUE ~ 0))

# Stats compared to team
# Add some stats to compare by team
players.2019 <- players.2019 %>% 
  group_by(tmID) %>% 
  mutate(
    tmPoints = points / sum(points),
    tmAssists = assists / sum(assists),
    tmRebounds = rebounds / sum(rebounds),
    tmDRebounds = case_when(
      dRebounds != 0 ~ dRebounds / sum(dRebounds),
      TRUE ~ 0),
    tmORebounds = case_when(
      oRebounds != 0 ~ oRebounds / sum(oRebounds),
      TRUE ~ 0)) %>% 
  ungroup()

# More individual stats
players.2019 <- players.2019 %>% 
  mutate(
    ftPct = ftMade / ftAttempted,
    fgPct = fgMade / fgAttempted,
    threePct = case_when(
      threeAttempted != 0 ~ threeMade / threeAttempted,
      TRUE ~ 0),
    efgPct = (fgMade + 0.5 * threeMade) / fgAttempted,
    astTovRatio = case_when( 
      turnovers != 0 ~ assists / turnovers,
      TRUE ~ 0), 
    dReboundPct = dRebounds / rebounds,
    oReboundPct = oRebounds / rebounds,
    totalGameScore = points + 0.4 * (fgMade + threeMade) - 0.7 * fgAttempted - 0.4 * (ftAttempted - ftMade) + 0.7 * oRebounds + 0.3 * dRebounds + steals + 0.7 * assists + 0.7 * blocks - 0.4 * PF - turnovers,
    avgGameScore = totalGameScore / GP)

players.2019
```

## Complete Player Data

The following code basically follows the previous process, but loops through the web pages for the 2013 through 2019 regular seasons in order to compile a complete data set.

```{r}
# all-nba data frame ------------------------------------------------------

allnba.url <- "https://en.wikipedia.org/wiki/All-NBA_Team"
webpage <- read_html(allnba.url) 

allnba <- webpage %>% 
  html_nodes(xpath = '//*[@id="mw-content-text"]/div/table[5]') %>% 
  html_table(fill = T)

# Extract only players from the relevant seasons
allnba <- allnba[[1]][,-c(3,5,7)] %>% 
  tail(30)

# Label the season consistently and rename column to 'year'
allnba$Season <- rep(2013:2018, each = 5)

# Gather data on team/player to create yearly list
allnba <- allnba %>% gather(Team, Player, -Season)

# nba data ----------------------------------------------------------------

years <- 2013:2019
players.2019 <- data.frame() # Instantiate data frame

for(year in years){
  # Store appropriate url in xml object
  url <- paste0("https://www.basketball-reference.com/leagues/NBA_", year, "_totals.html")
  webpage <- read_html(url)
  
  # Get headers
  headers <- webpage %>% 
    html_nodes("table#totals_stats > thead > tr > th") %>%
    html_text %>% 
    .[-1] # Omit the "Rank" column
  
  # Store data
  data <- webpage %>%
    html_nodes("table#totals_stats > tbody > tr > td") %>% 
    html_text %>% 
    matrix(ncol = length(headers), byrow = TRUE) %>%
    as.data.frame
  
  # Add headers and year
  names(data) <- headers
  data$year <- year
  
  # Correct data types and grab columns in the same order as the Kaggle data
  data <- data %>% 
    mutate_at(vars(G:PTS), as.character) %>% # convert to character before numeric (factor data type will cause issues going straight to numeric)
    mutate_at(vars(G:PTS), as.numeric) %>% 
    select(c(Player, year, Tm, G, MP, PTS, ORB, DRB, TRB, AST, STL, BLK, TOV, PF, FGA, FG, FTA, FT, "3PA", "3P", Pos))
  
  # Match column names to Kaggle data set
  names(data) <- c("playerID", "year", "tmID", "GP", "minutes", "points", "oRebounds", "dRebounds", "rebounds", "assists", "steals", 
                           "blocks", "turnovers", "PF", "fgAttempted", "fgMade", "ftAttempted", "ftMade", "threeAttempted", "threeMade", "pos")
  
  # Add position indicators
  data <- data %>% 
    mutate(
      center = case_when(
        grepl("C", pos) ~ as.integer(1),
        TRUE ~ as.integer(0)),
      
      forward = case_when(
        grepl("F", pos) ~ as.integer(1),
        TRUE ~ as.integer(0)),
      
      guard = case_when(
        grepl("G", pos) ~ as.integer(1),
        TRUE ~ as.integer(0))) %>% 
    select(-pos)
  
  # All-NBA Indicator
  if(year != 2019){
    
    # Get list to check against from main list at top
    allnba.check <- allnba %>% 
      filter(Season == year) %>% 
      .$Player %>% 
      gsub("(\\^)|\\s\\([0-9]+\\)$", "", .) # Remove extra characters from names
    
    data <- data %>%
      mutate(allNBA = case_when(
        playerID %in% allnba.check ~ as.integer(1),
        TRUE ~ as.integer(0)))
    
  } else {
    data$allNBA <- 0 
  }
  
  # All-Star Webpage
  allstar.url <- paste0("https://www.basketball-reference.com/allstar/NBA_", year, ".html")
  
  # Get All-Star roster
  allstar <- allstar.url %>% 
    read_html %>% 
    html_nodes("table > tbody > tr > th > a") %>% 
    html_text
  
  # Add allstar column
  data <- data %>% 
    mutate(allstar = case_when(
      playerID %in% allstar ~ as.integer(1),
      TRUE ~ as.integer(0)))
  
  # Create variables for per-game statistics
  data <- data %>% 
    mutate(
      minutesPerGame = minutes / GP,
      pointsPerGame = points / GP,
      assistsPerGame = assists / GP,
      oReboundsPerGame = oRebounds / GP,
      dReboundsPerGame = dRebounds / GP,
      reboundsPerGame = rebounds / GP,
      stealsPerGame = steals / GP,
      blocksPerGame = blocks / GP,
      turnoversPerGame = turnovers / GP,
      fgAttemptedPerGame = fgAttempted / GP,
      fgMadePerGame = fgMade/ GP,
      ftAttemptedPerGame = ftAttempted / GP,
      ftMadePerGame = ftMade / GP,
      threeAttemptedPerGame = threeAttempted / GP,
      threeMadePerGame = threeMade / GP)
  
  # GP Ratio, health, and stats compared to league 
  data <- data %>%
    mutate(
      GPRatio = GP / max(GP), 
      healthy = case_when(
        GPRatio >= 0.7 ~ as.integer(1),
        TRUE ~ as.integer(0)),
      lgPoints = points / sum(points),
      lgAssists = assists / sum(assists),
      lgRebounds = rebounds / sum(rebounds),
      lgDRebounds = case_when(
        dRebounds != 0 ~ dRebounds / sum(dRebounds),
        TRUE ~ 0),
      lgORebounds = case_when(
        oRebounds != 0 ~ oRebounds / sum(oRebounds),
        TRUE ~ 0))
  
  # Add some stats to compare by team
  data <- data %>% 
    group_by(tmID) %>% 
    mutate(
      tmPoints = points / sum(points),
      tmAssists = assists / sum(assists),
      tmRebounds = rebounds / sum(rebounds),
      tmDRebounds = case_when(
        dRebounds != 0 ~ dRebounds / sum(dRebounds),
        TRUE ~ 0),
      tmORebounds = case_when(
        oRebounds != 0 ~ oRebounds / sum(oRebounds),
        TRUE ~ 0)) %>% 
    ungroup()
  
  # More individual stats
  data <- data %>% 
    mutate(
      ftPct = case_when(
        ftAttempted != 0 ~ ftMade / ftAttempted,
        TRUE ~ 0),
      fgPct = case_when(
        fgAttempted != 0 ~ fgMade / fgAttempted,
        TRUE ~ 0),
      threePct = case_when(
        threeAttempted != 0 ~ threeMade / threeAttempted,
        TRUE ~ 0),
      efgPct = case_when(
        fgAttempted != 0 ~ (fgMade + 0.5 * threeMade) / fgAttempted,
        TRUE ~ 0),
      astTovRatio = case_when( 
        turnovers != 0 ~ assists / turnovers,
        TRUE ~ 0), 
      dReboundPct = case_when(
        rebounds != 0 ~ dRebounds / rebounds,
        TRUE ~ 0),
      oReboundPct = case_when(
        rebounds != 0 ~ oRebounds / rebounds,
        TRUE ~ 0),
      totalGameScore = points + 0.4 * (fgMade + threeMade) - 0.7 * fgAttempted - 0.4 * (ftAttempted - ftMade) + 0.7 * oRebounds + 0.3 * dRebounds + steals + 0.7 * assists + 0.7 * blocks - 0.4 * PF - turnovers,
      avgGameScore = case_when(
        GP != 0 ~ totalGameScore / GP,
        TRUE ~ 0))
  
  # Add to players df
  players.2019 <- rbind(players.2019, data)
}

write.csv(players.2019, "players_2019.csv", row.names = FALSE)

```

Now all we have to do is save the new data so we can use it with our model!

```{r save data}
write.csv(players.2019, "players_2019.csv", row.names = FALSE)
```

