---
title: "All-NBA Team Capstone"
author: "James Martinez"
date: "March 12, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
```

## Introduction

This capstone project will utilize historical NBA statistics from 1937 to 2012 to predict All-NBA Teams. Data used can be found [here](https://www.kaggle.com/open-source-sports/mens-professional-basketball).

## Loading the data sets

```{r}
# Store data as a tbl
players <- as_tibble(read.csv("basketball_players.csv"))
awards <- as_tibble(read.csv("basketball_awards_players.csv"))
allstar <- as_tibble(read.csv("basketball_player_allstar.csv"))
master <- as_tibble(read.csv("basketball_master.csv"))

# Check the players data
players
```

```{r, echo=FALSE}
players$playerID <- as.character(players$playerID)
awards$playerID <- as.character(awards$playerID)
allstar$player_id <- as.character(allstar$player_id)
```


Player data already appears to be tidy. As expected, nearly all of our variables of interest are numeric, and it looks like several variables (e.g. *oRebounds*, *dRebounds*, *threeAttempted*, *threeMade*) were not tracked in the early seasons. 


## Exploratory Data Analysis and Data Cleaning

First, the `setequal()` and `distinct()` functions are used to check for duplicate observations (just in case).

```{r}
# Check for duplicate observations
setequal(players, distinct(players))
setequal(awards, distinct(awards))
setequal(allstar, distinct(allstar))
```

### Trimming Some Data

#### The *players* Data Set

We'll be working with *players* first. The data set currently contains data from other leagues (e.g., ABA), as well as data from seasons before the first All-Star Game (pre-1950). We can start by removing these extra observations. Post-season stats, players with no games played, and miscellaneous features are also removed (Note that `lgID` is also removed after filtering).

```{r}
# Remove extraneous rows and columns 
players <- players %>% 
  filter(lgID == "NBA", year >= 1950, GP != 0) %>% 
  select(-contains("Post"), -c(note, stint, lgID, GS))

players
```



Now that the data set has been trimmed a bit, we can use the `summary()` function to view some general summary statistics.

```{r, echo=FALSE}
# Only summarize numeric variables
players %>% 
  select(-c(playerID, tmID)) %>% 
  summary()
```

Unfortunately, it looks like there are still several statistics that were not recorded since the inception of the All-Star game (which plenty of NBA historians would love to remind us of). We're concerned specifically with **rebounds and 3-pointers**. A bit of research reveals that [the NBA didn't begin recording offensive and defensive rebounds until the 1973-1974 season](https://en.wikipedia.org/wiki/Rebound_(basketball)), and [the 3-pointer wasn't adopted until 1979](https://www.usab.com/youth/news/2011/06/the-history-of-the-3-pointer.aspx). This means that we should only keep data from 1979 and later, but let's confirm that with `which.min()` just in case:

```{r}
# Group by year and sort by maxima
check <- players %>% 
  select(year, oRebounds, dRebounds, threeAttempted) %>% 
  group_by(year) %>% 
  summarize_if(is.numeric, max)

# Store the indices for the earliest year that each stat maxed out at 0
keep <- c(
  which.min(check$oRebounds == 0), 
  which.min(check$dRebounds == 0), 
  which.min(check$threeAttempted == 0))

# What's the earliest year we can keep?
check$year[max(keep)]
  
```

To prevent bias we need to retain data only for years in which **every statistic is available**, so we should get rid of observations before the 1979-1980 season.

```{r}
players <- players %>% 
  filter(year >= 1979)

players
```

Look's like we've shaved the *players* data set down from **23,751 x 42** to **14,577 x 20**!

#### The *awards* Data Set

Now we can make similar adjustments to the data sets for end-of-season awards and All-Star games, which are much smaller than the *players* data set:

```{r}
awards
allstar
```

Similar to *players*, the *awards* data set has data for non-NBA leagues. It also has awards other than the All-NBA teams, such as **"Most Valuable Player"** and **"Rookie of the Year."** We'll only keep NBA data and relevant awards (Note that `grepl()` is used here, along with an appropriate regular expression).

```{r}
# Filter for league and year
awards <- awards %>% 
  filter(lgID == "NBA", year >= 1979) %>% 
  select(-c(note, lgID))

# Keep relevant awards
keep <- grepl("All-NBA\\sFirst|All-NBA\\sSecond|All-D|Valuable|Defensive", awards$award)
awards <- awards[keep,]

# Check for duplicate entries
setequal(awards, distinct(awards))
```

To prevent potential bias, **"All-NBA Third Team"** honors are disregarded, since the Third Team has only been selected from 1988 and on. **"Defensive Player of the Year"** is retained as there are only 3 seasons for which data is missing (Defensive POTY has been awarded since 1982). The `pos` variable is also retained, as each team must consist of two guards, two forwards, and one center. 

As a final step, we create indicator variables for each award. However, the *awards* data set currently stores multiple rows for years in which a player won multiple awards. For instance:

```{r}
awards %>% filter(playerID == "abdulka01", year == 1979) %>% group_by(year)
```

To prevent any odd behavior when merging with the original data set, we need to combine such observations into a single entry. This is accomplished with `summarize(award = paste(award, collapse = ", "))` after grouping by playerID and year. `case_when()` and `grepl()` are then used to create the desired indicator variables.

```{r}
# Create indicator variables for each award
awards <- awards %>% 
  group_by(playerID, year) %>% 
  summarize(award = paste(award, collapse = ", ")) %>% 
  mutate(
    allDefFirstTeam = case_when(
      grepl("All-Defensive\\sFirst", award) ~ as.integer(1),
      TRUE ~ as.integer(0)),
    
    allDefSecondTeam = case_when(
      grepl("All-Defensive\\sSecond", award) ~ as.integer(1),
      TRUE ~ as.integer(0)),
    
    allNBAFirstTeam = case_when(
      grepl("All-NBA\\sFirst", award) ~ as.integer(1),
      TRUE ~ as.integer(0)),
    
    allNBASecondTeam = case_when(
      grepl("All-NBA\\sSecond", award) ~ as.integer(1),
      TRUE ~ as.integer(0)),
    
    MVP = case_when(
      grepl("Most", award) ~ as.integer(1),
      TRUE ~ as.integer(0)),
    
    defPOTY = case_when(
      grepl("Defensive\\sPlayer", x = award) ~ as.integer(1),
      TRUE ~ as.integer(0)))

awards
```

#### The *masters* Data Set

The *masters* data set is the master reference for first and last name data and, more importantly, the associated player ID. It also contains additional biographical data, such as height, weight, and college attended. The only relevant variable for us (other than player ID) is the position played, so we'll omit everything else.

```{r}
# Select relevant variables, rename, and create indicators
pos <- master %>% 
  select(bioID, pos) %>% 
  rename("playerID" = "bioID") %>% 
  mutate(
    center = case_when(
      grepl("C", pos) ~ as.integer(1),
      TRUE ~ as.integer(0)),
    
    forward = case_when(
      grepl("F", pos) ~ as.integer(1),
      TRUE ~ as.integer(0)),
    
    guard = case_when(
      grepl("G", pos) ~ as.integer(1),
      TRUE ~ as.integer(0))
  ) %>% 
  select(-pos)

pos$playerID <- as.character(pos$playerID) # Initialize as a character vector
pos
```

#### The *allstar* Data Set

We wish to use All-Star data as an additional explanatory variable. To avoid complications with existing player statistics, we can disregard actual game performance and simply create an indicator variable for team  membership.

```{r}
# Filter for league and year
allstar <- allstar %>% 
  filter(league_id == "NBA", season_id >= 1979) %>%
  select(player_id, season_id) %>%
  mutate(allstar = as.integer(1)) %>% 
  rename("playerID" = "player_id", "year" = "season_id") # Rename variables to match format of players

# Check for duplicate entries
setequal(allstar, distinct(allstar))
```

Oops, looks like there's a duplicate entry. After a prolonged investigation, we find that Marques Johnson was incorrectly coded as *johnsma02* instead of *johnsma01*, causing a discrepancy for the All-Star team in 1979, which had both Marques Johnson *and* Magic Johnson.

```{r echo=FALSE}
dummy <- as_tibble(read.csv("basketball_player_allstar.csv"))
dummy %>%
  filter(first_name == "Marques")
```

We can correct this quirk with `allstar$player_id[allstar$first_name == "Marques"] <- "johnsma01"` and re-run our code.

```{r echo=FALSE}
allstar <- as_tibble(read.csv("basketball_player_allstar.csv"))
allstar$player_id[allstar$first_name == "Marques"] <- "johnsma01"
allstar <- allstar %>% 
  filter(league_id == "NBA", season_id >= 1979) %>%
  select(player_id, season_id) %>%
  mutate(allstar = as.integer(1)) %>% 
  rename("playerID" = "player_id", "year" = "season_id") 
```

```{r}
# Check for duplicate entries
setequal(allstar, distinct(allstar))
```

Much better! Now we should be fine when we merge our data.

### Merging the Data

Now that we've cleaned up each data set a bit, we can `left_join()` all of our data into a single data set. `left_join()` takes two arguments, `x` and `y`, and augments the primary data frame (`x`) with the data from the secondary data frame (`y`). The `join` is performed based on a common key which should be present in both data frames (in our case, **playerID and year**). In the following code, we can join the data frames separately to check whether rows are added (which would indicate that the join didn't go as expected). If there is data from the secondary data frame 

```{r}
# Merge player data with awards data
players %>% 
  left_join(awards, by = c("playerID", "year"))

# Merge player data with position data
players %>% 
  left_join(pos, by = c("playerID"))

# Merge player data with allstar data
players %>% 
  left_join(allstar, by = c("playerID", "year"))
```

The *awards* and *pos* data sets seemed to merge correctly, but it looks like the number of rows changed after the *allstar* merge. To investigate this error, we can use the `anti_join()` function to compare the original *allstar* data set to the entries that were actually transferred over. `anti_join()` only returns rows from the primary data frame that are *not* in the secondary, while keeping the columns from the primary.

```{r}
anti_join(allstar, players)
```

After manually cross-referencing the excluded All-Star rows and playerID's with [Basketball-Reference](https://www.basketball-reference.com/allstar/), [Wikipedia](https://en.wikipedia.org/wiki/NBA_All-Star_Game), and the master playerID list, we find that all of the missing rows had their *playerID* values incorrectly coded per the *players* and *master* data sets, which resulted in skipped entries during the join (e.g., *o'neaje01*, *o'neasha01*). We can correct these errors and try joining the data again.

```{r, echo=FALSE}
allstar <- as_tibble(read.csv("basketball_player_allstar.csv"))
allstar$player_id <- as.character((allstar$player_id))
```

```{r}
# Fix coding errors for all players in the anti_join
allstar$player_id[allstar$first_name == "Marques" & allstar$last_name == "Johnson"] <- "johnsma01"
allstar$player_id[allstar$first_name == "Magic" & allstar$last_name == "Johnson"] <- "johnsma02"
allstar$player_id[allstar$first_name == "Kenny" & allstar$last_name == "Anderson"] <- "anderke01"
allstar$player_id[allstar$first_name == "Carmelo" & allstar$last_name == "Anthony"] <- "anthoca01"
allstar$player_id[allstar$first_name == "Eric" & allstar$last_name == "Floyd"] <- "floydsl01"
allstar$player_id[allstar$first_name == "Manu" & allstar$last_name == "Ginobili"] <- "ginobem01"
allstar$player_id[allstar$first_name == "Kevin" & allstar$last_name == "Johnson"] <- "johnske02"
allstar$player_id[allstar$first_name == "Larry" & allstar$last_name == "Johnson"] <- "johnsla02"
allstar$player_id[allstar$first_name == "Eddie" & allstar$last_name == "Jones"] <- "jonesed02"
allstar$player_id[allstar$first_name == "Lafayette" & allstar$last_name == "Lever"] <- "leverfa01"
allstar$player_id[allstar$first_name == "Rashard" & allstar$last_name == "Lewis"] <- "lewisra02"
allstar$player_id[allstar$first_name == "Jermaine" & allstar$last_name == "O'Neal"] <- "onealje01"
allstar$player_id[allstar$first_name == "Shaquille" & allstar$last_name == "O'Neal"] <- "onealsh01"
allstar$player_id[allstar$first_name == "Jim" & allstar$last_name == "Paxson"] <- "paxsoji02"
allstar$player_id[allstar$first_name == "Jeff" & allstar$last_name == "Rutland"] <- "rulanje01"
allstar$player_id[allstar$first_name == "Wally" & allstar$last_name == "Szczerbiak"] <- "szczewa02"
allstar$player_id[allstar$first_name == "David" & allstar$last_name == "Thompson"] <- "thompda01"
allstar$player_id[allstar$first_name == "Antoine" & allstar$last_name == "Walker"] <- "walkean02"
allstar$player_id[allstar$first_name == "Kevin" & allstar$last_name == "Willis"] <- "willike02"

# Apply previous filters, etc.
allstar <- allstar %>% 
  filter(league_id == "NBA", season_id >= 1979) %>%
  select(player_id, season_id) %>%
  mutate(allstar = as.integer(1)) %>% 
  rename("playerID" = "player_id", "year" = "season_id") 

# Retry the join and save into allstar.join
allstar.join <- players %>% 
  left_join(allstar, by = c("playerID", "year"))
allstar.join
```

The resulting data frame, *allstar.join*, has the correct amount of 14577 observations, which confirms that there are no extranneous rows added. As a quick aside, summing up the *allstar* column will show a number **greater than** 743:

```{r}
# Add all observations for which the player was an All-Star that year
sum(allstar.join$allstar, na.rm = TRUE)
```

The could indicate that some players were incorrectly coded as All-Stars. However, after a bit of digging with the `duplicated()` function, we find that these additional *allstar* entries are an artifact of the *players* structure; players who were traded during the season have **multiple rows for the year that they were traded**, so players who were traded that ALSO happened to make the All-Star team will have both rows coded with `allstar = 1`. 

```{r}
# Filter All-Stars and select rows to match the allstar data set
allstar.join <- allstar.join %>% 
  select(playerID, year, allstar) %>% 
  filter(allstar == 1)

# Check rows with duplicate entries
allstar.join[duplicated(allstar.join),]
```

The astute reader will notice that there are still 2 observations missing from the join (summing the *allstar* column resulted in 751 rows, but there are 10 additional rows added from traded All-Star players). 

```{r}
anti_join(allstar, allstar.join, by = c("playerID", "year"))
```

The line of code `allstar$player_id[allstar$first_name == "David" & allstar$last_name == "Thompson"] <- "thompda01"` should have corrected David Thompson's 1982 All-Star game appearance. But, after reviewing the original *allstar* data (again), we find that there were *two* entries for the *thomsda01* player ID: one for "David Thompson" and one for "David *Thomspon*". This is an error that won't affect our *players* data. The second observation (for Magic Johnson) is for the [1992 All-Star game](https://en.wikipedia.org/wiki/1992_NBA_All-Star_Game). From Wikipedia, 

> The game is most remembered for the return of Los Angeles Lakers legend Magic Johnson, who retired before the 1991-92 NBA season after contracting HIV.

With this, we've taken care of the many discrepancies that arose from the attempted joining of the *allstar* and *player* data. At this point we can be confident that we can safely merge these data sets with *players*.

```{r}
# Join all data sets
players.joined <- players %>% 
  left_join(pos, by = "playerID") %>% 
  left_join(awards, by = c("playerID", "year")) %>% 
  left_join(allstar, by = c("playerID", "year")) %>% 
  mutate_if(is.numeric, coalesce, 0L) # Return non-All-Star entries as integer zeroes

players.joined
```

Now all that's left is to save the cleaned data set!

``` {r}
write.csv(players.joined, "players_clean.csv", row.names = FALSE)
```
